Structure Learning for $\VAR(p)$ Models

### 2019-05-17 paper/paper.tex
Shifted around various sections and removed some superfluous statements.

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage[backend=bibtex,sorting=none]{biblatex}
\usepackage{enumitem}
\bibliography{\string~/Documents/academics/global_academics/global_bib.bib}
\usepackage{hyperref}

@@

We will work with a discrete time and wide-sense stationary process
$x(t)$ taking elements in $L_2$.  We suppose that $x(t)$ has zero
mean, $\E x(t) = 0$, and the matrix valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$ is absolutely summable and has a spectra $S(\omega)$:

@@

\begin{equation}
  \label{eq:hilbert_space_defn}
  \begin{aligned}
    \H^x &= \cl \{\sum_{\tau = -T}^T a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, T \in \N\} \subseteq L_2(\Omega),\\
    \H_t^x &= \cl \{\sum_{\tau = 0}^T a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, T \in \N\} \subseteq L_2(\Omega),
  \end{aligned}
\end{equation}

where the closure is naturally in mean-square.  $\H_t^x$ is the space
spanned by the values of $x(t)$ between the ``infinite past'' and
``now'', we will often omit the superscript $x$ which should be clear
from context.  Evidently these spaces are separable and as closed
subspaces of a Hilbert space they are themselves Hilbert.  We will
denote the spaces generated in analagous ways by particular components
of $x$ as e.g. $\H_t^{(i, j)}$, $\H_t^{i}$ or by all but particular
components as e.g. $\H_t^{-j}$.

@@

\begin{theorem}[Granger Causality Equivalences]
  \label{thm:granger_causality_equivalences}
  Let $x(t)$ be a WSS process with absolutely summable covariance
  sequence, and spectral density uniformly bounded above and below.  Denote by
  $\xi_{ij} \defeq \linEerr{x_i(t)}{\H^{-j}}$ and
  $\xi_i \defeq \linEerr{x_i(t)}{\H_t}$.  Then, the following are equivalent:

  \begin{enumerate}
    \item{$x_j \ngc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$}
    \item{$\mathcal{F}_{j \rightarrow i} \defeq \ln\frac{\xi_i}{\xi_{ij}} = 0$}
    \item{$\H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j}$ \hl{Modify to look just at $H_t$}}
    \item{$\linE{x_i(t)}{\H_{t - 1}^{-j}} = \linE{x_i(t)}{\H_{t - 1}}$}
    % \item{\hl{Wold $A(\tau)$ condition}}
    % \item{Tie in the 0s of $S(\omega)^{-1}$, though this isn't immediately directed.}
  \end{enumerate}
\end{theorem}

Statement (4) means that
$(y - \linE{y}{\H_{t - 1}^{-j}}) \perp (z - \linE{z}{\H_{t -
    1}^{-j}})\ \forall y \in \H_t^i,\ z \in \H_{t - 1}^j$, which is an
appealing geometric statement that there is no direct feedback from
$x_j$ to $x_i$ \cite{lindquist}.

@@

@@

$(2) \Rightarrow (4)$: In computing $(y - \linE{y}{\H_{t = 1}^{-j}})$ for
$y \in \H_t^i$ it is sufficient to consider $y = x_i(t)$ since
$\H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$, in which case
$(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$.  $(4)$ follows
since $v_i(t) \perp \H_{t - 1}$ and
$\forall z \in \H_{t - 1}^j\ (z - \linE{z}{\H_{t - 1}^{-j}}) \in \H_{t
  - 1}$

\hl{TODO: $(4) \Rightarrow (1)$}

\end{proof}

@@

\begin{definition}[Pairwise Granger-causality]
  We will say that $x_j$ pairwise Granger-causes $x_i$ and write
  $x_j \pwgc x_i$ if $x_i$ Granger-causes $x_j$ with respect only to
  $(x_i, x_j)$.
\end{definition}

This notion is of interest for a variety of reasons.  From a purely
theoretical or conceptual standpoint, we will seek to understand when
and how pairwise causality can capture the idea of ``flow of
information'' in the underlying graph.  It will also be useful to
reason about what the conditions under which \textit{unobserved}
components of $x(t)$ may or may not interfere with inference in the
actually observed components.  Finally, we will seek to construct
practical estimation procedures based purely on pairwise causality
tests and understand their properties.

We first need to establish some graph theoretic notation and
terminology, collected formally in definitions for the reader's
convenient reference.

@@

@@

\begin{equation}
  \sum_{\tau = 1}^\infty B_{ij}(\tau)x_j(t - \tau) \defeq \B_{ij}(z)x_j(t).
\end{equation}

@@

We continue to develop the relationship between $x_i$ and it's family tree.

@@

  where $\wtB_{ii}(z) = (1 - \B_{ii}(z))^{-1}$ and
  $\wtB_{ik}(z) = \wtB_{ii}(z) B_{ik}(z)$.  \hl{Is this true: }$\wtB_{ii} = \A_{ii}$ ?

@@

This statement is ultimately about the sparsity pattern in the Wold
decomposition matrices $A(\tau)$ since
$x_i(t) = \sum_{\tau = 0}^\infty \sum_{j = 1}^n a_{ij}(\tau)v_j(t -
\tau)$.  The proposition states that $\A_{ij}(\tau) = 0$ unless
$j \in \anc{i}$.

@@

\begin{proof}
  From equation (\ref{eqn:ar_representation}), which we are assuming
  throughout the paper to be invertible, we can write

@@

  \begin{equation}
    \label{eqn:resolvant_inv}
    (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k.
  \end{equation}

  Now, let $S$ be a $0-1$ matrix containing the sparsity pattern of
  $\B(z)$.  Then from lemma \ref{lem:adj_matrix} $S$ is the transpose
  of the adjacency matrix for $\gcg$ and hence $(S^k)_{ij}$ is
  non-zero if and only if $j \in \gpn{k}{i}$, and therefore
  $\B(z)^k_{ij} = 0$ if $j \not \in \gpn{k}{i}$.  The Cayley-Hamilton
  theorem can then be applied to represent the infinite sum of equation
  (\ref{eqn:resolvant_inv}) in terms only of \textit{finite} powers of
  $\B(z)$, so the application of lemma \ref{lem:adj_matrix} in the
  limit is not necessary.  Finally, since $\anc{i} = \cup_{k = 1}^n\gpn{k}{i}$
  we see that $\A(z)_{ij}$ is zero if $j \not\in \anc{i}$.

@@

Before moving onto our next proposition, we need two lemmas formalizing
some intuitively apparent facts.

\begin{lemma}
  \label{lem:subspace_sum_projection}
  Let $x \in \H$ where $\H$ is a Hilbert space having inner product $\inner{\cdot}{\cdot}$ and let $\{\H_n\}_{n = 0}^\infty$ be a collection of subspaces of $\H$ such that $x \perp \H_0$.  Then

  \begin{equation*}
    \linE{x}{\bigvee_{n = 0}^\infty\H_n} = \linE{x}{\bigvee_{n = 1}^\infty\H_n},
  \end{equation*}

  where $\H_1 \vee \H_2 = \cl \{\alpha + \beta\ |\ \alpha \in \H_1, \beta \in \H_2 \}$ is the closed sum of subspaces.
\end{lemma}
\begin{proof}
  We can apply the Gram-schmidt process to the collection of subspaces to obtain a sequence of orthogonal subspaces $\widetilde{\H}_0, \widetilde{\H}_1, \ldots$ with $\widetilde{\H}_0 = \H_0$, $\widetilde{\H}_n \subseteq \H_n$, and $\bigvee_{n = 0}^\infty\H_n = \bigoplus_{n = 0}^\infty \widetilde{\H}_n$, where $\oplus$ indicates the orthogonal sum of subspaces (which is always closed).  Then, we have an explicit representation of the projection

  \begin{align*}
    \linE{x}{\bigvee_{n = 0}^\infty\H_n} &= \sum_{n = 0}^\infty \inner{x}{x_n} x_n\\
    &\overset{(a)}{=} \sum_{n = 1}^\infty \inner{x}{x_n} x_n,
  \end{align*}

  where $x_n \in \widetilde{\H}_n \subseteq \H_n$ and $(a)$ follows because of our assumption that $x \perp \H_0$.  This establishes that the projection is infact a member of the smaller subspace $\bigvee_{n = 1}^\infty\H_n$, and since the error can only increase by projecting onto this smaller subspace, we have the conclusion.
\end{proof}

The following lemma is reproduced from \cite{lindquist} here for convenience.

\begin{lemma}[\cite{lindquist}]
  \label{lem:conditional_orthogonality_equivalence}
  Consider three closed subspaces of a Hilbert space $\mathcal{A}$, $\mathcal{B}$, $\mathcal{X}$.  The following statements are equivalent

  \begin{enumerate}
    \item{$\mathcal{A} \perp \mathcal{B}\ |\ \mathcal{X}$}
    \item{$\linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}}\ \forall \beta \in \mathcal{B}$}
    % \item{$\linE{\beta}{\mathcal{A}} = \linE{\linE{\beta}{\mathcal{X}}}{\mathcal{A}}\ \forall \beta \in \mathcal{B}}$}
  \end{enumerate}
\end{lemma}
\begin{proof}
  Proceding from the definition of conditional orthogonality we have

  \begin{align*}
    &\inner{\alpha - \linE{\alpha}{\mathcal{X}}}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
    &{\iff} \inner{\alpha + x - \linE{\alpha + x}{\mathcal{X}}}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
    &\overset{(a)}{\iff} \inner{\alpha + x}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
    &\overset{(b)}{\iff} \linE{\beta - \linE{\beta}{\mathcal{X}}}{\mathcal{A} \vee \mathcal{X}} = 0\ \forall \beta \in \mathcal{B}\\
    &\overset{(c)}{\iff} \linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}},
  \end{align*}

  where $(a)$ follows since $(\beta - \linE{\beta}{\mathcal{X}}) \perp \mathcal{X}$ and $\linE{\alpha + x}{\mathcal{X}} \in \mathcal{X}$, $(b)$ since $z \perp \mathcal{X} \vee \mathcal{A} \iff \linE{z}{\mathcal{X} \vee \mathcal{A}} = 0$, and $(d)$ since $\linE{z}{\H_2} = \linE{\linE{z}{\H_2}}{\H_1}$ for $\H_2 \subseteq \H_1$.

  % This is property (vi) in Lindquist
  % \begin{align*}
  %   \langle \alpha - \linE{\alpha}{\mathcal{X}}, \beta - \linE{\beta}{\mathcal{X}} \rangle &= 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
  %   \overset{(a)}{\iff} \langle \alpha, \beta - \linE{\beta}{\mathcal{X}} \rangle &= 0\ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
  %   \overset{(b)}{\iff} \linE{\beta - \linE{\beta}{\mathcal{X}}}{\mathcal{A}} &= 0\ \forall \beta \in \mathcal{B}
  % \end{align*}

  % where $(a)$ follows because $\linE{\alpha}{\mathcal{X}} \in \mathcal{X}$ and $\beta - \linE{\beta}{\mathcal{X}} \in \mathcal{X}$, and $(b)$ follows since $\beta \perp \mathcal{A} \iff \linE{\beta}{\mathcal{A}} = 0$.
\end{proof}

\begin{proposition}
  \label{prop:ancestor_properties}
  Consider two distinct nodes $i, j$ of a Granger-causality graph
  $\gcg$.  Then, if

@@

\begin{remark}
  Note that $H_t^{(i)} \perp \H_t^{(j)}$ states that the
  \textit{present} of $x_i(t)$ is uncorrelated with the \textit{past}
  of $x_j(t)$, but we need not have $\H_t^{(i)} \perp \H_t^{(j)}$
  since $i \in \anc{j}$ is not excluded.

@@

The previous proposition can be strengthened significantly for the
case of Granger-causality.  Note that it is possible to have some
$k \in \anc{i} \cap \anc{j}$ where still $j \npwgc i$, for example
consider the three node graph $k \rightarrow i \rightarrow j$ where
clearly $k \in \anc{i}\cap\anc{j}$ but $j \npwgc i$.  We must
introduce the concept of a \textit{confounding} variable, which
effectively eliminates the possibility presented in this example.

@@

\begin{proposition}
  If in a Granger-causality graph $\gcg$ we have $j \pwgc i$ then
  $j \in \anc{i}$ or $\exists k \in \anc{i} \cap\anc{j}$ which is a
  confounder of $(i, j)$.
\end{proposition}

@@

  In order to deploy a recursive argument, define the following
  partition of $\pa{u}$, for some node $u$: (\hl{TODO: Draw a figure depicting these sets})

@@

  which follows from Proposition \ref{prop:ancestor_properties}.
  Finally, for $k \in C_1(j)$ the case $k = i$ is immediate (since the
  error in estimating $\Phi(z)\B_{ii}(z)x_i(t)$ given
  $\H_{t - 1}^{(i)}$ is $0$), so suppose $k \ne i$.  We know from
  above that $k$ inherits the key properties referred to as $a, b$ and
  therefore we can recursively expand $k$ in the same way as in
  equation (\ref{eqn:xj_partition_expansion}).  Continuing this
  recursion for each $k \in C_1(j)$ (where $k \ne i$) must eventually
  terminate since $i \in \anc{k}$.
\end{proof}

@@

\begin{corollary}
  If the graph $\gcg$ is acyclic and if $j \pwgc i$ and $i \pwgc j$ then $\anc{i} \cap \anc{j} \ne \emptyset$.
\end{corollary}

It seems intuitive that a converse of proposition \ref{prop:ancestor_properties}
would hold, i.e. $j \in \anc{i} \Rightarrow j \pwgc i$.  Unfortunately,
this is not the case in general, as different paths through $\gcg$ can
lead to cancellation (see example \ref{ex:diamond_cancellation}).  In
fact, we do not even have $j \in \pa{i} \Rightarrow j \pwgc i$ (see
example \ref{ex:lag_cancellation}).

@@

In this section and the next we will seek to understand when converse
statements of Proposition \ref{prop:ancestor_properties} \textit{do}
hold.  One possibility is to restrict to coefficients of the system
matrix matrix, e.g. by requiring that $B_{ij}(\tau) \ge 0$.  Instead,
we think it is more meaningful to focus on the defining feature of
time series networks, that is, the topology of $\gcg$.

@@

Examples of strongly causal graphs include directed trees (or
forests), DAGs where each node has at most one parent, and figure
\ref{fig:example_fig3} of this paper.  The persistence condition is
used to ensure that $\linE{x_i(t)}{\H_{t - 1}^{(i)}} \ne 0$ i.e. that
the each process has at least some amount of ``memory'' -- this can
occur either due to cycles, or non-zero $\B_{ii}(z)$ terms in equation
\eqref{eqn:ar_representation}.  It is evident that the strong causal property
is inherited by subgraphs, but persistence need not be, unless $\gcg$ is a DAG.

\hl{Not sure these are needed, but were important in my own thinking.}

\begin{lemma}
  \label{lem:still_strongly_causal}
  If $\gcg$ is a strongly causal graph, any subgraph formed by eliminating nodes
  as well as all of the in and out edges thereof is still strongly causal.
\end{lemma}
\begin{proof}
  If $\gcg$ has at most one path between any two nodes, there can only
  be fewer paths after removing nodes from $\gcg$.
\end{proof}

% \begin{lemma}
%   \label{lem:still_persistent}
%   If $\gcg$ is a persistent DAG, any subgraph formed by eliminating nodes
%   as well as all of the in and out edges thereof is still persistent.
% \end{lemma}
% \begin{proof}
%   \hl{TODO -- is this needed?}
% \end{proof}

We explore a number of properties of strongly causal graphs before
moving into the main result of section \ref{sec:theory} in section
\ref{sec:pairwise_algorithm}.  The following important property
essentially strengthens proposition \ref{prop:ancestor_properties}
for the case of strongly causal graphs.

@@

  and therefore that $H_t^{(i)} \not\perp\ \H_{t - 1}^{(j)}\ |\ \H_{t - 1}^{(i)}$, which by theorem (\ref{thm:granger_causality_equivalences}) and \hl{lemma [name the lemma]} is enough to establish that $j \pwgc i$.

@@

\begin{remark}
  It can be seen from the conclusion of the proof that simple facts
  about Granger causality are indeed reliant on our assumptions about
  the nature of $x(t)$ laid out in the section \ref{sec:theory}, in
  particular, the innovations process $v(t)$ must be full rank.

  Moreover, that strong conditions need to be placed on the topology
  of $\gcg$ in order for proposition \ref{prop:pwgc_anc} to hold can
  be seen through the earlier examples.
\end{remark}

\begin{lemma}
  If $\gcg$ is strongly causal, then any $\VAR(p)$ system on $\gcg$ is stable if and only if
  $B_{ii}(z)$ is stable for every $i = 1, \ldots, n$.
\end{lemma}
\begin{proof}
  \hl{This is certainly true, but I don't think I need to use it anywhere.}
\end{proof}

\begin{remark}
  \hl{Possibly expand on this thought}

>>>>

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage[backend=bibtex,sorting=none]{biblatex}
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\bibliography{\string~/Documents/academics/global_academics/global_bib.bib}
\usepackage{hyperref}

@@

We will work with a discrete time and wide-sense stationary process
$x(t)$ taking elements in $L_2$.  We suppose that $x(t)$ (with
$t \in \Z$) has zero mean, $\E x(t) = 0$, and the matrix valued
covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$ is absolutely
summable and has a spectra $S(\omega)$:

@@

\begin{equation}
  \label{eq:hilbert_space_defn}
  \begin{aligned}
    % \H^x &= \cl \{\sum_{\tau = -T}^T a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, T \in \N\} \subseteq L_2(\Omega),\\
    \H_t^x &= \cl \{\sum_{\tau = 0}^T a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, T \in \N\} \subseteq L_2(\Omega)\\
    H_t^x &= \{a x(t)\ |\ a \in \R\} \subseteq L_2(\Omega),
  \end{aligned}
\end{equation}

where the closure is naturally in mean-square.  $\H_t^x$ is the space
spanned by the values of $x(t)$ between the ``infinite past'' and
``now'', we will often omit the superscript $x$ which should be clear
from context.  Evidently these spaces are separable, and as closed
subspaces of a Hilbert space they are themselves Hilbert.  We will
denote the spaces generated in analagous ways by particular components
of $x$ as e.g. $\H_t^{(i, j)}$, $\H_t^{i}$ or by all but particular
components as e.g. $\H_t^{-j}$.

@@

We require two technical facts we refer to in the sequel, the notion
of conditional orthogonality is used throughout.

\begin{lemma}[\cite{lindquist} Proposition 2.4.2]
  \label{lem:conditional_orthogonality_equivalence}
  Consider three closed subspaces of a Hilbert space $\mathcal{A}$,
  $\mathcal{B}$, $\mathcal{X}$.  The following statements are
  equivalent

  \begin{enumerate}
    \item{$\mathcal{A} \perp \mathcal{B}\ |\ \mathcal{X}$}
    \item{$\linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}}\ \forall \beta \in \mathcal{B}.$}
    % \item{$\linE{\beta}{\mathcal{A}} = \linE{\linE{\beta}{\mathcal{X}}}{\mathcal{A}}\ \forall \beta \in \mathcal{B}}$}
    \end{enumerate}

    Where $\mathcal{A} \perp \mathcal{B}\ |\ \mathcal{X}$ denotes
    conditional orthogonality:

    \begin{equation*}
      \inner{a - \linE{a}{\mathcal{X}}}{b - \linE{b}{\mathcal{X}}} = 0\ \forall a \in \mathcal{A}, b \in \mathcal{B}.
    \end{equation*}
\end{lemma}
% \begin{proof}
%   Proceding from the definition of conditional orthogonality we have

%   \begin{align*}
%     &\inner{\alpha - \linE{\alpha}{\mathcal{X}}}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
%     &{\iff} \inner{\alpha + x - \linE{\alpha + x}{\mathcal{X}}}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
%     &\overset{(a)}{\iff} \inner{\alpha + x}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
%     &\overset{(b)}{\iff} \linE{\beta - \linE{\beta}{\mathcal{X}}}{\mathcal{A} \vee \mathcal{X}} = 0\ \forall \beta \in \mathcal{B}\\
%     &\overset{(c)}{\iff} \linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}},
%   \end{align*}

%   where $(a)$ follows since $(\beta - \linE{\beta}{\mathcal{X}}) \perp \mathcal{X}$ and $\linE{\alpha + x}{\mathcal{X}} \in \mathcal{X}$, $(b)$ since $z \perp \mathcal{X} \vee \mathcal{A} \iff \linE{z}{\mathcal{X} \vee \mathcal{A}} = 0$, and $(d)$ since $\linE{z}{\H_2} = \linE{\linE{z}{\H_2}}{\H_1}$ for $\H_2 \subseteq \H_1$.

%   % This is property (vi) in Lindquist
%   % \begin{align*}
%   %   \langle \alpha - \linE{\alpha}{\mathcal{X}}, \beta - \linE{\beta}{\mathcal{X}} \rangle &= 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
%   %   \overset{(a)}{\iff} \langle \alpha, \beta - \linE{\beta}{\mathcal{X}} \rangle &= 0\ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
%   %   \overset{(b)}{\iff} \linE{\beta - \linE{\beta}{\mathcal{X}}}{\mathcal{A}} &= 0\ \forall \beta \in \mathcal{B}
%   % \end{align*}

%   % where $(a)$ follows because $\linE{\alpha}{\mathcal{X}} \in \mathcal{X}$ and $\beta - \linE{\beta}{\mathcal{X}} \in \mathcal{X}$, and $(b)$ follows since $\beta \perp \mathcal{A} \iff \linE{\beta}{\mathcal{A}} = 0$.
% \end{proof}
\begin{lemma}
  \label{lem:subspace_sum_projection}
  Let $x \in \H$ where $\H$ is a separable Hilbert space having inner product
  $\inner{\cdot}{\cdot}$ and let $\{\H_n\}_{n = 0}^N$ be a
  collection of subspaces of $\H$ such that $x \perp \H_0$.  Then

  \begin{equation*}
    \linE{x}{\bigvee_{n = 0}^N\H_n} = \linE{x}{\bigvee_{n = 1}^N\H_n},
  \end{equation*}

  where
  $\H_1 \vee \H_2 = \cl \{\alpha + \beta\ |\ \alpha \in \H_1, \beta
  \in \H_2 \}$ is the closed sum of subspaces.
\end{lemma}
\begin{proof}
  Apply the Gram-schmidt process and directly calculate the projections.
\end{proof}

\begin{theorem}[Granger Causality Equivalences]
  \label{thm:granger_causality_equivalences}
  Let $x(t)$ be a WSS process with absolutely summable covariance
  sequence, and spectral density uniformly bounded above and below.
  Denote by $\xi_{ij} \defeq \linEerr{x_i(t)}{\H^{-j}}$ and
  $\xi_i \defeq \linEerr{x_i(t)}{\H_t}$ for distinct $i, j$.  Then,
  the following are equivalent:

  \begin{enumerate}
    \item{$x_j \ngc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$ i.e. $\B_{ij}(z) = 0$}
    \item{$\mathcal{F}_{j \rightarrow i} \defeq \ln\frac{\xi_i}{\xi_{ij}} = 0$}
    \item{$\H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j} \iff H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j}$}
    \item{$\linE{x_i(t)}{\H_{t - 1}^{-j}} = \linE{x_i(t)}{\H_{t - 1}}$}
    % \item{\hl{Wold $A(\tau)$ condition}}
    % \item{Tie in the 0s of $S(\omega)^{-1}$, though this isn't immediately directed.}
  \end{enumerate}
\end{theorem}

@@

  We have $(4) \iff (5)$ as a result of Lemma
  \ref{lem:conditional_orthogonality_equivalence}, the additional
  equivalence in $(4)$ is apparent since $\H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$.

@@

$(2) \Rightarrow (4)$: In computing $(y - \linE{y}{\H_{t - 1}^{-j}})$ for
$y \in H_t^i$ it is sufficient to consider $y = x_i(t)$ by linearity, then since
$H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$ we have
$(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$ since $\B_{ij}(z) = 0$.  $(4)$ then follows
since $v_i(t) \perp \H_{t - 1}$ and
$\forall z \in \H_{t - 1}^j\ (z - \linE{z}{\H_{t - 1}^{-j}}) \in \H_{t
  - 1}$
\end{proof}

@@

\subsection{Granger Causality Graphs}
We first need to establish some graph theoretic notation and
terminology, collected formally in definitions for the reader's
convenient reference.

@@

Our principle object of study will be a graph determined by
Granger-causality relations as follows.

@@

\begin{equation}
  \B_{ij}(z)x_j(t) \defeq \sum_{\tau = 1}^\infty B_{ij}(\tau)x_j(t - \tau).
\end{equation}

@@

We continue to develop the relationship between $x_i$ and it's family
tree, where the following proposition is analagous to the definitions
of feedback free processes of \cite{caines1975feedback}.

@@

  where $\wtB_{ii}(z) = (1 - \B_{ii}(z))^{-1}$ and
  $\wtB_{ik}(z) = \wtB_{ii}(z) B_{ik}(z)$.

@@

This statement is ultimately about the sparsity pattern in the Wold
decomposition matrices $A(\tau)$ since
$x_i(t) = \sum_{\tau = 0}^\infty \sum_{j = 1}^n A_{ij}(\tau)v_j(t -
\tau)$.  The proposition states that $\A_{ij}(z) = 0$ unless
$j \in \anc{i}$.

@@

\begin{proof}
  Equation \eqref{eqn:parent_expansion} was derived prior to the
  statement of the theorem.  We are left to demonstrate equation
  \eqref{eqn:ancestor_expansion}.

  From equation (\ref{eqn:ar_representation}), which we are assuming
  throughout the paper to be invertible, we can write

@@

  \begin{equation}
    \label{eqn:resolvant_inv}
    (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k,
  \end{equation}

  where we can apply the Cayley-Hamilton theorem to write the infinite
  sum of \eqref{eqn:resolvant_inv} in terms of \textit{finite} powers
  of $\B$.

  Let $S$ be a $0-1$ matrix containing the sparsity pattern of
  $\B(z)$, from lemma \ref{lem:adj_matrix} $S$ is the transpose of the
  adjacency matrix for $\gcg$ and hence $(S^k)_{ij}$ is non-zero if
  and only if $j \in \gpn{k}{i}$, and therefore $\B(z)^k_{ij} = 0$ if
  $j \not \in \gpn{k}{i}$.  Finally, since
  $\anc{i} = \bigcup_{k = 1}^n\gpn{k}{i}$ we see that $\A_{ij}(z)$ is
  zero if $j \not\in \anc{i}$.

@@

\begin{definition}[Pairwise Granger-causality]
  We will say that $x_j$ pairwise Granger-causes $x_i$ and write
  $x_j \pwgc x_i$ if $x_j$ Granger-causes $x_i$ with respect only to
  $(x_i, x_j)$.
\end{definition}

This notion is of interest for a variety of reasons.  From a purely
theoretical or conceptual standpoint, we will seek to understand when
and how pairwise causality can capture the idea of ``flow of
information'' in the underlying graph.  It may also be useful for
reasoning about what the conditions under which \textit{unobserved}
components of $x(t)$ may or may not interfere with inference in the
actually observed components.  Finally, motivated from a practical
standpoint to analyze causation in large systems, we will seek to
construct practical estimation procedures based purely on pairwise
causality tests which can result in reduced computational complexity.

\begin{proposition}
  \label{prop:ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger-causality graph
  $\gcg$.  Then, if

@@

\begin{remark}
  Note that $H_t^{(i)} \perp \H_t^{(j)}$ states that the
  \textit{present} of $x_i(t)$ is uncorrelated with $x_j(t)$,
  but we need not have $\H_t^{(i)} \perp \H_t^{(j)}$ since
  $i \in \anc{j}$ is not excluded.

@@

The previous proposition can be strengthened significantly for the
case of Granger-causality.  Note that it is possible to have some
$k \in \anc{i} \cap \anc{j}$ where still $j \npwgc i$, consider for example
the three node graph $k \rightarrow i \rightarrow j$ where
clearly $k \in \anc{i}\cap\anc{j}$ but $j \npwgc i$.  We must
introduce the concept of a \textit{confounding} variable, which
effectively eliminates the possibility presented in this example.

@@

\begin{proposition}
  \label{prop:ancestor_properties}
  If in a Granger-causality graph $\gcg$ we have $j \pwgc i$ then
  $j \in \anc{i}$ or $\exists k \in \anc{i} \cap\anc{j}$ which is a
  confounder of $(i, j)$.
\end{proposition}

@@

  In order to deploy a recursive argument, define the following
  partition of $\pa{u}$, for some node $u$:

@@

  which follows from Proposition \ref{prop:ancestor_uncorrelated}.
  Finally, for $k \in C_1(j)$ the case $k = i$ is immediate (since the
  error in estimating $\Phi(z)\B_{ii}(z)x_i(t)$ given
  $\H_{t - 1}^{(i)}$ is $0$), so suppose $k \ne i$.  We know from
  above that $k$ inherits the key properties referred to as $a, b$ and
  therefore we can recursively expand $k$ in the same way as in
  equation (\ref{eqn:xj_partition_expansion}).  Continuing this
  recursion for each $k \in C_1(j)$ (where $k \ne i$) must eventually
  terminate since $i \in \anc{k}$.
\end{proof}

@@

\begin{corollary}
  If the graph $\gcg$ is acyclic and if $j \pwgc i$ and $i \pwgc j$ then $\exists k \in \anc{i} \cap \anc{j}$ confounding $(i, j)$.
\end{corollary}

It seems reasonable to expect a converse of proposition
\ref{prop:ancestor_properties} to hold, i.e.
$j \in \anc{i} \Rightarrow j \pwgc i$.  Unfortunately, this is not the
case in general, as different paths through $\gcg$ can lead to
cancellation (see example \ref{ex:diamond_cancellation}).  In fact, we
do not even have $j \in \pa{i} \Rightarrow j \pwgc i$ (see example
\ref{ex:lag_cancellation}).

@@

In this section and the next we will seek to understand when converse
statements of Proposition \ref{prop:ancestor_properties} \textit{do}
hold.  One possibility is to restrict the coefficients of the system
matrix, e.g. by requiring that $B_{ij}(\tau) \ge 0$.  Instead,
we think it is more meaningful to focus on the defining feature of
time series networks, that is, the topology of $\gcg$.

@@

Examples of strongly causal graphs include directed trees (or
forests), DAGs where each node has at most one parent, and figure
\ref{fig:example_fig3} of this paper.  It is evident that the strong
causal property is inherited by subgraphs.

% \begin{lemma}
%   \label{lem:still_strongly_causal}
%   If $\gcg$ is a strongly causal graph, any subgraph formed by eliminating nodes
%   as well as all of the in and out edges thereof is still strongly causal.
% \end{lemma}
% \begin{proof}
%   If $\gcg$ has at most one path between any two nodes, there can only
%   be fewer paths after removing nodes from $\gcg$.
% \end{proof}

For later use, and to get a feel for the topological implications of
strong causality We explore a number of properties of such graphs
before moving into the main result of this section.  The following
important property essentially strengthens proposition
\ref{prop:ancestor_properties} for the case of strongly causal graphs.

@@

  and therefore that $H_t^{(i)} \not\perp\ \H_{t - 1}^{(j)}\ |\ \H_{t - 1}^{(i)}$, which by theorem (\ref{thm:granger_causality_equivalences}) is enough to establish that $j \pwgc i$.

@@

% \begin{remark}
%   It can be seen from the conclusion of the proof that simple facts
%   about Granger causality are indeed reliant on our assumptions about
%   the nature of $x(t)$ laid out in the section \ref{sec:theory}, in
%   particular, the innovations process $v(t)$ must be full rank.

%   Moreover, that strong conditions need to be placed on the topology
%   of $\gcg$ in order for proposition \ref{prop:pwgc_anc} to hold can
%   be seen through the earlier examples.
% \end{remark}

% This holds in any DAG
% \begin{lemma}
%   If $\gcg$ is strongly causal, then any $\VAR(p)$ system on $\gcg$ is stable if and only if
%   $B_{ii}(z)$ is stable for every $i = 1, \ldots, n$.
% \end{lemma}
% \begin{proof}
%   \hl{This is certainly true, but I don't think I need to use it anywhere.}
% \end{proof}

\begin{remark}
  As we have seen and as is true in much of statistics, confounding
  nodes pose challenges for Granger-causality.  However, as opposed to
  Pearl's causal calculus \cite{pearl2000art}, pairwise
  Granger-causality does not suffer any difficulty with so-called
  ``colliders'', that is, the topology $i \rightarrow k \leftarrow j$
  will never result in $i \pwgc j$ or $j \pwgc i$.  This is evidently
  an advantage of the \textit{temporal} nature of Granger causality --
  ``information'' cannot flow backwards along the edges of $\gcg$.
\end{remark}
